Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-08 manuscript/manuscript.tex
additions

\vb{The presence of an unselected control line provides an alternative way to
partition the effects of linked selection and genetic drift: we can compare the
total variance in allele frequency change of the control line (which excludes
the effect of artificial selection on allele frequencies) to the total variance
in frequency change of the Longshanks selection lines. We estimate at least
32\% (95\% CI $[21\%, 48\%]$) of the variance in allele frequency change is
driven by the effects of selection, of which 14\% (95\% CI $[3\%, 33\%]$) is
estimated to be unique to a selection line, and 17\% (95\% CI $[9\%, 23\%]$) is
the effect of shared selection between the two Longshanks selection lines.}

@@

Since the seminal analysis of \textcite{Maynard_Smith1974-lc} demonstrating
that linked neutral diversity is reduced as an advantageous polymorphism arises
and sweeps to fixation, over four decades of theoretical and empirical research
has enriched our understanding of linked selection.  One under-used approach to
understand the genome-wide effects of selection on standing variation, e.g.
selection on an infinitesimal polygenic trait, stems from an early quantitative
genetic model of linked selection \parencite{Robertson1961-ho} and its later
developments
(\cite{Santiago1995-hx,Santiago1998-bs,Wray1990-zf,J_A_Woolliams_N_R_Wray_R_Thompson2008-qo};
\gc{THE WRAY PAPER DATE IS WRONG}
see also \cite{Barton2000-zg} for a comparison of these models with classic
hitchhiking models). Implicit in these models is that autocovariance between
allele frequency change is created when there is heritable fitness variation in
the population, a signal that may be readily detected from temporal genomic
data \parencite{Buffalo2019-io}.  Depending on how many loci affect fitness,
such a strong effect of linked selection may not be differentiable from genetic
drift using only single contemporary population samples or looking at at
temporal allele frequency change at each locus in isolation.  In this way,
averaging summaries of temporal data allows us to sidestep the key problem of
detecting selection from standing variation: that the genomic footprint leaves
too soft of a signature to differentiate from a background of genetic drift. In
fact we find that the temporal covariance signal is detectable even in the most
extremely difficult to detect soft sweep case: polygenic selection on highly
polygenic traits \parencite{Buffalo2019-io}.

@@

Our analyses reveal that a sizable proportion of allele frequency change in
these populations is due to the (indirect) action of selection. Capitalizing on
replicated designs, we characterized the extent to which convergent selection
pressures lead to parallel changes in allele frequencies across replicate
populations, and found that a reasonable proportion of the response is shared
across short timescales. These likely represent substantial under-estimates of
the contribution of linked selection because \vb{the studies we have reanalyzed
do not sequence the population each generation, preventing us from including
the effects of stronger correlations between adjacent generations.
Furthermore, \gc{our} estimation methods are intentionally
conservative,\gc{for example they exclude the contribution of selection that
  does not persist across generations and selection that reverses sign
  thus they show be seen as a storng lower bound of the effects of 
selection.}

@@

Our primary focus here has been on evolution in laboratory populations. It is
unclear whether we should expect a similar impact of selection in natural
populations. In some of these experiments, selection pressures may have been
stronger or more sustained that in natural populations
\vb{\parencite{Hendry1999-zu,Hairston2005-ga}}. Conversely, these lab
populations were maintained at very small population sizes, small effective
population sizes 300, 450, and 45 for the \textcite{Kelly2019-dc}, and
\textcite{Castro2019-uk} \gc{MISSING REF??} studies respectively, which will amplify the role of
genetic drift. The advantage of lab experiments is that they are closed
populations, in natural populations temporal covariance could also arise from
the systematic migration of alleles from differentiated populations.
\vb{Adapting} these methods to natural populations will require either
populations that are reasonably closed to migration, or for the effect of
migration to be accounted for possibly either by knowledge of allele
frequencies in source populations or the identification of migrant individuals.

@@

\begin{align}
  \Delta p_{t,\mathrm{LS1}} &= \Delta_{_D} p_{t,\mathrm{LS1}} + \Delta_{_{U}} p_{t,\mathrm{LS1}} + \Delta_{_\mathrm{LS}} p_t \\ 
  \Delta p_{t,\mathrm{LS2}} &= \Delta_{_D} p_{t,\mathrm{LS2}} + \Delta_{_{U}} p_{t,\mathrm{LS2}} + \Delta_{_\mathrm{LS}} p_t \\ 
  \Delta p_{t,\mathrm{Ctrl}} &= \Delta_{_D} p_{t,\mathrm{Ctrl}} + \Delta_{_{U}} p_{t,\mathrm{Ctrl}}
\end{align}

\begin{align}
  (\var(\Delta p_{t,\mathrm{LS1}}) + \var(\Delta p_{t,\mathrm{LS2}}))/2 - \var(\Delta p_{t,\mathrm{Ctrl}}) = 
  \overline{\var(\Delta_{_D} p_{t,\mathrm{LS}})} + \overline{\var(\Delta_{_U} p_{t,\mathrm{LS}})} + \var(\Delta_{_\mathrm{LS}} p_t)
\end{align}

where the bars indicate values averaged

We note that there very well may be selection acting in the control line...

We can partition the allele frequency change between the two timepoints (20
generations) for a Longshanks line as \vb{$\Delta p_{t,\mathrm{LS1}} =
\Delta_{_{D}} p_{t,\mathrm{LS1}} + \Delta_{_{U}} p_{t,\mathrm{LS1}} +
\Delta_{_S} p_{t,\mathrm{LS}}$} where these  terms in are the decomposition are
the allele frequency change due to drift in \vb{Longshanks replicate 1
  ($\Delta_{_D} p_{t,\mathrm{LS1}}$), selection unique to the LS1 replicate
  ($\Delta_{_U} p_{t,\mathrm{LS1}}$), and selection response shared between the
two Longshanks replicates ($\Delta_{_S} p_{t,\mathrm{LS}}$) respectively (and
similarly for the Longshanks two line, LS2)}. By construction we will assume
that each of these terms are uncorrelated within replicates, and that only the
shared term covaries between the replicates. Assuming that we can approximate
the contribution of genetic drift in the LS lines the variance in allele
frequency change observed in the control, i.e. \vb{$\var(\Delta
p_{t,\mathrm{Ctrl}}) = \var(\Delta_{_{D}}
p_{t,\mathrm{LS2}})=\var(\Delta_{_{D}} p_{t,\mathrm{LS2}})$}, then we can
estimate the increase in variance in allele frequency change due to selection
as $(\var(\Delta p_{t,\mathrm{LS1}}) + \var(\Delta p_{t,\mathrm{LS2}}))/2 -
\var(\Delta p_{t,\mathrm{Ctrl}})$ and the shared effect of selection across
selected lines as $\cov(\Delta p_{t,\mathrm{LS1}}, \Delta p_{t,\mathrm{LS2}})$.

>>>>

\vb{The presence of an unselected control line provides an alternative way to
  partition the effects of linked selection and genetic drift: we can compare
  the total variance in allele frequency change of the control line (which
  excludes the effect of artificial selection on allele frequencies) to the
  total variance in frequency change of the Longshanks selection lines. We can
  partition the allele frequency change between the two timepoints (20
  generations) for a Longshanks line as \vb{$\Delta p_{t,\mathrm{LS1}} =
    \Delta_{_{D}} p_{t,\mathrm{LS1}} + \Delta_{_{U}} p_{t,\mathrm{LS1}} +
    \Delta_{_S} p_{t,\mathrm{LS}}$} where these  terms in are the decomposition
    are the allele frequency change due to drift in \vb{Longshanks replicate 1
      ($\Delta_{_D} p_{t,\mathrm{LS1}}$), selection unique to the LS1 replicate
      ($\Delta_{_U} p_{t,\mathrm{LS1}}$), and selection response shared between
      the two Longshanks replicates ($\Delta_{_S} p_{t,\mathrm{LS}}$)
      respectively (and similarly for the Longshanks two line, LS2)}. By
      construction we will assume that each of these terms are uncorrelated
      within replicates, and that only the shared term covaries between the
      replicates. Assuming that we can approximate the contribution of genetic
      drift in the LS lines the variance in allele frequency change observed in
      the control, i.e. \vb{$\var(\Delta p_{t,\mathrm{Ctrl}}) =
      \var(\Delta_{_{D}} p_{t,\mathrm{LS2}})=\var(\Delta_{_{D}}
    p_{t,\mathrm{LS2}})$}, then we can estimate the increase in variance in
    allele frequency change due to selection as $(\var(\Delta
    p_{t,\mathrm{LS1}}) + \var(\Delta p_{t,\mathrm{LS2}}))/2 - \var(\Delta
    p_{t,\mathrm{Ctrl}})$ and the shared effect of selection across selected
    lines as $\cov(\Delta p_{t,\mathrm{LS1}}, \Delta p_{t,\mathrm{LS2}})$.  We
    estimate at least 32\% (95\% CI $[21\%, 48\%]$) of the variance in allele
    frequency change is driven by the effects of selection, of which 14\% (95\%
  CI $[3\%, 33\%]$) is estimated to be unique to a selection line, and 17\%
(95\% CI $[9\%, 23\%]$) is the effect of shared selection between the two
Longshanks selection lines.}

@@

Since the seminal analysis of \textcite{Maynard_Smith1974-lc} demonstrating
that linked neutral diversity is reduced as an advantageous polymorphism arises
and sweeps to fixation, over four decades of theoretical and empirical research
has enriched our understanding of linked selection.  One under-used approach to
understand the genome-wide effects of selection on standing variation, e.g.
selection on an infinitesimal polygenic trait, stems from an early quantitative
genetic model of linked selection \parencite{Robertson1961-ho} and its later
developments
(\cite{Santiago1995-hx,Santiago1998-bs,Wray1990-zf,Woolliams1993-qo}; see also
\cite{Barton2000-zg} for a comparison of these models with classic hitchhiking
models). Implicit in these models is that autocovariance between allele
frequency change is created when there is heritable fitness variation in the
population, a signal that may be readily detected from temporal genomic data
\parencite{Buffalo2019-io}.  Depending on how many loci affect fitness, such a
strong effect of linked selection may not be differentiable from genetic drift
using only single contemporary population samples or looking at at temporal
allele frequency change at each locus in isolation.  In this way, averaging
summaries of temporal data allows us to sidestep the key problem of detecting
selection from standing variation: that the genomic footprint leaves too soft
of a signature to differentiate from a background of genetic drift. In fact we
find that the temporal covariance signal is detectable even in the most
extremely difficult to detect soft sweep case: polygenic selection on highly
polygenic traits \parencite{Buffalo2019-io}.

@@

Our analyses reveal that a sizable proportion of allele frequency change in
these populations is due to the (indirect) action of selection. Capitalizing on
replicated designs, we characterized the extent to which convergent selection
pressures lead to parallel changes in allele frequencies across replicate
populations, and found that a reasonable proportion of the response is shared
across short timescales. These likely represent substantial under-estimates of
the contribution of linked selection because the studies we have reanalyzed do
not sequence the population each generation, preventing us from including the
effects of stronger correlations between adjacent generations.  Furthermore,
our estimation methods are intentionally conservative, for example they exclude
the contribution of selection that does not persist across generations and
selection that reverses sign thus they show be seen as a strong lower bound of
the effects of selection.

@@

Our primary focus here has been on evolution in laboratory populations. It is
unclear whether we should expect a similar impact of selection in natural
populations. In some of these experiments, selection pressures may have been
stronger or more sustained that in natural populations
\vb{\parencite{Hendry1999-zu,Hairston2005-ga}}. Conversely, these lab
populations were maintained at very small population sizes, small effective
population sizes 300, 450, and 45 for the \textcite{Barghi2019-qy},
\textcite{Kelly2019-dc}, and \textcite{Castro2019-uk} studies respectively,
which will amplify the role of genetic drift. The advantage of lab experiments
is that they are closed populations, in natural populations temporal covariance
could also arise from the systematic migration of alleles from differentiated
populations.  \vb{Adapting} these methods to natural populations will require
either populations that are reasonably closed to migration, or for the effect
of migration to be accounted for possibly either by knowledge of allele
frequencies in source populations or the identification of migrant individuals.

@@

\begin{align}
  \Delta p_{t,\mathrm{LS1}} &= \Delta_{_D} p_{t,\mathrm{LS1}} + \Delta_{_{U}} p_{t,\mathrm{LS1}} + \Delta_{_\mathrm{LS}} p_t \\ 
  \Delta p_{t,\mathrm{LS2}} &= \Delta_{_D} p_{t,\mathrm{LS2}} + \Delta_{_{U}} p_{t,\mathrm{LS2}} + \Delta_{_\mathrm{LS}} p_t \\ 
  \Delta p_{t,\mathrm{Ctrl}} &= \Delta_{_D} p_{t,\mathrm{Ctrl}}
\end{align}

Note that since the control replicate does not undergo artificial selection, we
assume that its allele frequency changes are determined entirely by genetic
drift. With free mating individuals (such as in a cage population), this may
not be the case, and sequencing adjacent generations would allow one to more
easily differentiate the effects of selection and drift.

We assume that we can approximate the contribution of genetic drift in the
Longshanks selection lines with the observed variance in the control line, or
$\var(\Delta p_{t,\mathrm{Ctrl}}) = \var(\Delta_{_{D} p_{t,\mathrm{LS1}}}) =
\var(\Delta_{_{D} p_{t,\mathrm{LS2}}})$. Then, the combined effects of
selection can be estimated by averaging the variances of the two Longshanks
selection lines, and subtracting the variance in allele frequency change in the
control line, which we treat as driven by drift alone (since matings are
random). Note that each variance is bias-corrected according to the methods
described in Supplementary Materials \ref{supp:matrix-correction}, and the
average sequencing depths between lines are nearly identical. Thus, we have

\begin{align}
  (\var(\Delta p_{t,\mathrm{LS1}}) + \var(\Delta p_{t,\mathrm{LS2}}))/2 - \var(\Delta p_{t,\mathrm{Ctrl}}) = 
  \overline{\var(\Delta_{_U} p_{t,\mathrm{LS}})} + \var(\Delta_{_\mathrm{LS}} p_t)
\end{align}

where the bar indicate values averaged both Longshanks selection lines.
Additionally, use the fact that

\begin{align}
  \cov(\Delta p_{t,\mathrm{LS1}}, \Delta p_{t,\mathrm{LS2}}) = \var(\Delta_{_\mathrm{LS}} p_t)
\end{align}

we can also separate out the unique and shared components by subtracting off
this covariance,

\begin{align}
  \overline{\var(\Delta_{_U} p_{t,\mathrm{LS}})}  = 
(\var(\Delta p_{t,\mathrm{LS1}}) + \var(\Delta p_{t,\mathrm{LS2}}))/2 - \var(\Delta p_{t,\mathrm{Ctrl}}) - \cov(\Delta p_{t,\mathrm{LS1}}, \Delta p_{t,\mathrm{LS2}}).
\end{align}

Finally, we can divide each of these values by the total variance to get the
proportion of total variance drift, and unique and shared effects of selection
contribute towards the total. To derive confidence intervals for the estimates
of unique and shared effects of selection, we use a block bootstrap procedure
as described in Supplementary Materials Section \ref{supp:block-bootstrap}.
